# Theoretical framwork and political stability in Switzerland

## Social blocs, political equilibrium and systemic crisis

The empirical work conducted here and in the following sections are extensively based on the neorealist approach developed by Amable and Palombarini [-@amable_leconomie_2005; -@amable_neorealist_2008]. The point of departure of this approach is the fundamental diversity of social expectations coming from the heterogeneity of the social structure. This diversity leads to a wide variety of social interests which are in perpetual contradiction and discord. Societies are thus characterized by permanent social conflict, which cannot be forever resolved but only temporarily regulated through the interaction between three spheres of social regulations: ideology, institutions and political mediation [@amable_multidimensional_2023].

Common socio-economic characteristics define socio-economic groups, which only partially translate into socio-political groups. The latter are groups which gather agents sharing common social expectation regarding public policies and the desired socio-economic model [@amable_search_2021]. Regarding the relationships between socio-economic and socio-political groups, neorealism avoids the pitfalls of both economic determinism and Idealism. On the one hand, socio-demographic characteristics and thus socio-economic groups are captured by socio-political groups: agents with, for instance, low income level have more incentive to be in favour of pro-redistribution policies. However, they may not be in favor of redistribution if they are culturally influenced by the dominant ideology to not follow their economic interests. Ideology thus mediates the translation of economic interests into explicit social demand. On the other hand, political mediation represents all the political strategies aimed at gathering socio-political groups into social blocs, which are thus almost never homogeneous. 

Finally, social blocs compete in the political arena with the objective to influence public policy towards the satisfaction of their social demands. A social bloc whose main demands are met and favored by public decision-making is the *dominant social bloc* (DBS). Social blocs whose demands are not favored by public policy are excluded from the DBS. In the neorealist framework, a political crisis corresponds to the collapse or destabilization of the DBS. Conversely, a stable existence of a DBS corresponds to a situation of *political equilibrium*. If the break-up of the DBS persists in time and no political strategy is successful in aggregating a new DBS, the crisis becomes *systemic* [@amable_diversity_2003; @amable_leconomie_2005; @amable_neorealist_2008; @amable_multidimensional_2023]. The concept of political equilibrium should not let the reader think that the existence of a DBS leads to an unchanging stable situation; and, conversely, periods of systemic crises can persist in time and last several decades, as in France [@amable_structural_2017]. The contrasts between Italy and France and the one side, and Switzerland on the other makes the latter particularly interesting for the neorealist approach. 

## Switzerland: Political Equilibrium and the Swiss Bourgeois Bloc 

In fact, relative to Italy and France, which underwent deep periods of political and systemic crises in the last decades, Switzerland's political stability stands rather remarkably. This stability should, at first sight, not constitute a surprise: Switzerland ranks consistently top among the best performing economies in a wide variety of socio-economic indicators: GDP per capita, Human Living Index (HDI), or even the Economic Complexity Index (ECI)[^intro-2]. If one considers confidence in government as a good indicator of political stability, and if one looks at the data provided by the OECD on that matter, Switzerland indeed stands out as one of the most stable countries in the OECD.

```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(haven)
library(quarto)
library(sjlabelled)
library(readr)
library(readxl)

```

```{r oecd trust}
#| warning: false
#| message: false
#| echo: false
#| fig-cap: "Share of people who report having confidence in the national government"
#| label: fig-trust1

# OECD (2023), "Trust in government" (indicator), https://doi.org/10.1787/1de9675e-en (accessed on 08 October 2023)
# https://www.oecd-ilibrary.org/governance/trust-in-government/indicator/english_1de9675e-en 

oecd_trust_gov <- read_csv("data/oecd_trust_gov.csv")


oecd_trust_gov %>% 
  filter(LOCATION %in% c("CHE", "FRA", "DEU", "ITA", "SWE", "GBR", "USA")) %>% 
  ggplot()+
  aes(x = TIME, y = Value, color = LOCATION, shape = LOCATION)+
  geom_line()+
  geom_point()+
  theme_bw()+
  scale_x_continuous(breaks = 2006:2022)+
  scale_color_brewer(palette = "Set1")+
  labs(title = "Political Stability across OECD countries",
       subtitle = "Trust in National Government: 2006-2022",
       y = "%", x = "",
       caption = "Source: OECD")+
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))

```

@fig-trust1 shows political stability as measured by the share of respondents who declared having trust in national government in a sample of OECD countries from 2006 to 2022. Switzerland shows the highest level of trust among the countries shown in the figure. Swiss trust in government shows an upward trend since 2009 and the share of Swiss respondents declaring trust in government is consistently equal or above 80% since 2016. This stands in sharp contrast with Italy, France or the US, which show low level of trust. The latter can be interpreted as a reflection of the persistent political crisis in which France and Italy are plunged. 

The origins of the Italian crisis goes back to 1992 after the break-up of the dominant social bloc (DBS). The latter, constituted by an alliance between small and large firms, the public sector, and classes dependent on transfers and rent (the employees of the large industries being excluded from the DBS) was made possible and sustained by the growth of the public debt which temporarily satisfied the heterogeneous demand of the DBS, but collapsed due to internal contradictions and public debt reduction policy. Various strategies conducted by Italian politicians tried to consolidate a new DBS, but with mixed results [@palombarini_rupture_2001; @amable_leconomie_2012].

In France, the crisis started with the instability of the traditional left and right blocs, which were subject to internal contradictions. European integration divided the left bloc in half, while right-wing political strategies were unable to satisfy both its popular and bourgeois elements. The crisis was partially resolved with the successful strategy of Macron to gather both the bourgeois groups of the left and right blocs in a new DBS, the *bourgeois bloc*, which is nonetheless unstable and fragile because of its restricted size [@amable_illusion_2018].

However, other data sources show a more contrasted history of Switzerland's political stability. The *VOX* studies, which conducted post-ballot surveys after each referendum voting session, were harmonized in a cumulative dataset which offer long-run data from 1981 to 2016. A similar figure as @fig-trust1 can thus be made to have a precise focus on Switzerland:



```{r import voxit}
#| warning: false
#| message: false
#| echo: false
voxit <- read_sav("data/voxit.sav")

```

```{r voxit table trust}
#| warning: false
#| message: false
#| echo: false
voxit %>% 
  group_by(annee) %>% 
  count(a22) %>% 
  drop_na() %>% 
  mutate(prop = n/sum(n),
         n_sample = sum(n)) %>% 
  ungroup() %>% 
  filter(a22 == 1) %>% 
  mutate(se = sqrt(prop*(1-prop)/n),
         lower = prop - qnorm(.975)*se,
         upper = prop + qnorm(.975)*se) -> trust_gov

```

```{r voxit plot trust}
#| warning: false
#| message: false
#| echo: false
#| fig-cap: "Share of Swiss respondent trusting the government: 1981-2016"
#| label: fig-trust2

trust_gov %>% 
  ggplot()+
  aes(x = factor(annee), y = prop, label = scales::percent(round(prop, 4)))+ 
  geom_col(fill = "darkred", alpha = 0.8)+
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), expand = c(0, 0))+
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.3, position = position_dodge(width = .9))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 50, vjust = 1, hjust=1),
        text = element_text(size = 11))+
  xlab("")+
  ylab("")+
  labs(title = "Confidence in Government",
       subtitle = "Switzerland, 1981-2016",
       caption = "Source: VoxIt")
```

@fig-trust2 shows that Switzerland underwent contrasted periods of political stability and crises in the last decades. A first period ranging from 1981 to 1987 was rather stable despite some variation in trust. @fig-trust2 also depicts well the crisis of the 90s, which constitutes the most turbulent period in Switzerland's contemporary history. The crisis of the 1990s can be seen with the fall in trust from 65% in 1987 to a historical low level of 39% of Swiss respondents who declared to not have confidence in the federal government in 1995, three years after the rejection of the referendum for the adhesion to the European Economic Area (EEC). Swiss journalists, political observers, activist and scholars wrote extensively on this period which constitutes a rare case of deep political and economic crisis. @brunner_crise_1997 show that Swiss citizens lost confidence in almost all political institutions and organizations during the 1990s. José Ribeaud, famous journalist who was one of the first Swiss television presenters, published the dramatically untitled book *When Switzerland will disappear* (*Quand la Suisse Disparaîtra*) in 1998. Journalists like José Ribeaud or Pietro Boschetti generally emphasize the multiple political scandals that happened during the 1990s. The first scandal was the resignation of the Radical Elisabeth Kopp from the federal council, in which she was the very first woman member, after she secretly tried to protect her husband from judiciary investigations. The "secret files scandal" happened the same year: the mass system of surveillance of the federal authorities was revealed to the Swiss public. The "secret files scandal" revealed that one twentieth of Swiss citizen and one third of foreign residents were recorded in the public authorities' files. The Eizenstat report (1997) and the Bergier commission (1996) destabilized further the coutry by revealing the sinister role of the Swiss government and banks during World War II [@ribeaud_quand_1998; @boschetti_conquete_2007].

Since the break-up of the post-war keynesian "historical compromise" between the Swiss Socialist Party (SSP) and the traditional right parties during the so-called "decade of all dangers" of the 90s [@boschetti_conquete_2007], the neoliberal transformation of the HSM was also followed by important political change, that some authors interpreted as various forms of latent political crises. The Swiss political scene is still marked by the rise of the Swiss People Party (SVP), one of the most successful far-right populist parties in Europe. The electoral success of the SVP is associated with the destabilization of the traditional right bloc, which entered a phase of "cacophonic" crisis [@meuwly_liberte_2008] which still persists in the early 2020s. 

In fact, all the Swiss traditional main parties, the Swiss Socialist Party (SSP) The Liberals (PLR) and the Christian Democratic People's Party of Switzerland (PDC), are in decline since the post-war period. On the right, the crisis can be seen through the increase in party merges and party re-configurations. In 2008, the Bourgeois Democratic Party (PBD) was created after a scission from the SVP. One year later, the Radical-Democratic party merged with the liberals to create the FDP. The PBD, whose creation disturbed the composition of the reputed unchangeable federal council, then merged with the PDC to create *Le Centre* in 2021. The latter could lead to major political change in the 2023 federal elections since the polls predict that this party could surpass the PLR in vote share and thus take another seat in the federal council, at the expense of the PLR, which created the Swiss political system and used to dominate Swiss politics in the 19th and 20th century[^3].

```{r}
#| warning: false
#| message: false
#| echo: false
#| include: false
party_share <- read_excel(path = "C:/Users/celal/OneDrive/Bureau/master-thesis/data/force_parti_ofs.xlsx",
                          sheet = 2)
party_share[party_share == "*"] <- NA
party_share <- data.frame(lapply(party_share, as.numeric))

party_share_long <- 
party_share  %>% 
  pivot_longer(cols = !`...1`,
               names_to = "party",
               values_to = "Party_strength") %>% 
  rename(year = "...1")

party_share_long %>% 
  filter(party %in% c("PLR..PRD..3.", "PDC", "PS", "UDC", "PES")) %>% 
  ggplot()+
  aes(x = year, y = Party_strength, color = party, shape = party)+
  geom_line()+
  geom_point()+
  theme_bw()+
  scale_color_brewer(palette = "Set1")

```


[^intro-2]: In 2018, Switzerland ranks third in terms of GDP per capita corrected for purchasing power parity among OECD countries [@oecd_oecd_2019, p.14]. Switzerland had also the highest HDI in 2021 [@nations_human_2022, p.272] and the second highest ECI in 2021 [@noauthor_country_nodate].

[^3]: See @noauthor_elections_2023