# Social blocs, socio-economic groups and political divides in Switzerland

```{r set up}
#| echo: false
#| warning: false
#| message: false

library(irr)
library(JuliaCall)
library(JuliaConnectoR)
library(tidyverse)
library(haven)
library(quarto)
library(sjlabelled)
library(readr)
library(readxl)
library(haven)
library(viridis)
library(stargazer)
library(gtsummary)
library(gt)
library(brms)
library(ggrepel)
library(cowplot)
library(ggpubr)
library(lme4)
library(prediction)
library(marginaleffects)
library(ggeffects)
library(ordinal)
library(broom)
library(knitr)
library(readr)
library(gtExtras)
library(broom.mixed)
library(GGally)
library(questionr)
library(car)
library(effects)
library(sjPlot)
library(margins)
library(cAIC4)
library(ordinal)
library(pacman)
library(easystats)
```

## The traditional Swiss bourgeois and left blocs: historical perspectives

A striking feature of the history of Switzerland's social blocs is the persistence and resilience of a liberal-conservative bloc since the creation of modern Switzerland in 1848, a bloc that the existing literature on the topic calls the *bourgeois bloc*. Tracing the history of Switzerland's social blocs is therefore essentially a matter of tracing that of the bourgeois bloc. However, we shall see that tracing this history calls into question the extent to which, from a neorealist perspective, this bourgeois bloc can truly be described as bourgeois.

The history of the Swiss bourgeois bloc goes back to the creation of modern Switzerland. The combination of a religious and urban-rural divide between Radicals-Protestants modernizers and Catholic federalists, which was historically one of the most structuring conflicts in Switzerland, led to the Sonderbund civil war (1847) and the foundation of modern Switzerland in 1848. The federal institutions elaborated during and in the decades following the first Swiss constitution of 1848 can be interpreted as a social compromise between the losers of the Sonderbund civil war, the Catholic-Conservative coalition opposed to the centralization and modernization of the Swiss Confederation, and the victorious Radicals who were on average Protestants and bourgeois from urban areas [@sciarini_politique_2023]. However, these two blocs were highly heterogeneous and subject to internal contradictions.

From the outset of their movement, the Radicals were gathered by an unitary illusion: the pretention to transcend cleavages running through Swiss society and unite all its elements thanks to an encompassing goal: the modernization and centralization of the Swiss state. Far from homogenous, the Radical bloc was a kaleidoscopic mass-movement constituted by the working class through the Grütli Union and by an elite gathering entrepreneurs, bankers, journalists, lawyers and teachers, who were nonetheless divided over the role of the Swiss state. The Democrats were partisans of a more centralized and interventionist state while the liberal or "manchesterian" Radicals, were more federalist and in favor of laissez-faire. This divide between interventionists and liberals was made explicit as early as 1852 over the issue of the nationalization of the Swiss railway industry. Supporters of a state-run railroad network, represented by for instance Jakob Stämplfi and Johann Jakob Speiser, were opposed by liberal Radicals such as Alfred Escher, who were industrialists and bankers close to the private railway companies, when they weren't the owners themselves [@meuwly_les_2010].

Regarding the Catholic-Conservative bloc, the latter was from the start a cross-class coalition, constituted by, of course, the Catholic elite, but also by peasants, farmers, artisans and the commercial middle class from rural areas. These groups were united by Catholicism, but also by anti-modernism and federalism. This conservative coalition was the first to make extensive use of the constitutional and optional referendum (respectively 1848 and 1874) to block the modernization promulgated by the Radicals. While the Radicals were convinced that the idea of the Swiss nation would transcend divisions and social conflict, the Catholic-Conservative believed that religion, rather than the Swiss state, would endorse this role [@altermatt_conservatism_1979].

As industrialization, proletarianization and unionization intensified in the late 19th, the hegemony of the Radical mosaic gradually shattered as the contradictions between the policy expectation of the popular and bourgeois bases increased. Part of the democrats and of the working class joined or even founded socialist organizations, as seen through the creation of the Swiss socialist party in 1888 by a former Radical, Albert Steck. The creation of the latter and of the Radical-Democratic party (PRD) in 1894 marked the beginning of the Radical shift towards liberal stances, which became clear-cut after the general strike of 1912, after which "a long-denied class conflict became evident" [@meuwly_les_2010, p.49]. Conversely, industrialization and class conflict represented an opportunity for an extension of the social basis of the Catholic-Conservatives bloc towards the working class. The global development of a Catholic social doctrine, beginning with the encyclical *Rerum Novarum* (1881), was integrated by the Swiss Catholic-Conservatives, who then tried to attract parts of the working class [@altermatt_conservatism_1979].

This polarization of class conflict around a left and right bloc became even more salient after the second general strike of 1918, which was the biggest strike and social protest in Swiss history. The latter was followed by the introduction of the proportional representation system: the Radicals lost their hegemony in the parliament and the Catholic-Conservatives, who meanwhile became the Christian-Democrats, and the Socialists became important political actors in the National Council. The Socialist Party (PS) consolidated a left bloc while the Radical-Democrats (PRD), the Christian-Democrats (PDC) and the Agrarians formed an alliance against the left. This coalition between the PRD, PDC and Agrarians and its close links with business association such as the Vorort (1870) led many observers to name this dominant coalition the "bourgeois bloc" [@mach_redefinition_2007].

However, the extent to which this dominant bloc can truly be described as bourgeois must be questioned. In the sense given by the literature [@mach_redefinition_2007], the bourgeois bloc refers to a political coalition between the Radical-Democrats, the Christian-Democrats and the Agrarians (now UDC) and its close ties with business interest associations. This bloc is hence bourgeois only because it gathers right-wing parties and business interest associations, a definition which is rather narrow and can be misleading since it does not take into account the social basis of the bloc. Calling the right-wing coalition bourgeois ignores, for instance, the fact that this bloc gathered some popular socio-economic groups such as small-business owners and peasants, mainly through the Agrarian Party. In the sense given by Amable and Palombarini [@amable_illusion_2017], the bloc bourgeois observed in France corresponds to the aggregation of the wealthiest and most educated groups of the traditional left and right blocs. I argue that, from a neorealist perspective, the Radical bloc of the 19th century and the right-wing coalition of the 20th cannot be described as bourgeois and that other terms such as "liberal-conservative" or simply "right-wing" bloc are more suited.

In fact, until the post-war period, the evolution of the Swiss social blocs can thus be resumed as follows. From the middle to the late 19th century, the Radicals formed a strongly heterogeneous but dominant social bloc. This dominant social bloc was composed by the working class (mainly through the Grütli Union), and an urban elite composed by the most educated (journalists, lawyers, teachers) and the wealthy industrial bourgeoisie (entrepreneurs, bankers...). This DSB was united by their willingness to modernize the Confederation against the Catholic-Conservatives. A second period, from the late 19th to the second World War, marked the recomposition of the DSB and its transformation into a liberal-conservative political bloc. Part of the working class and of the democrats (educated elite) joined the emerging socialist-left bloc while the Christian-Democrats attracted part of the working class while being progressively incorporated into the DSB, along with the Agrarian party which represented the peasant and farmer classes.

Analyses of the first Swiss post-electoral surveys which were conducted in the 1970s show the heterogeneity of the Swiss "bourgeois bloc" and that the word bourgeois is not valid for the right bloc as a whole. The social basis of the Liberal-Radical party was constituted by large employers/liberal professions and managers. The UDC relied heavily on small business owners' support while the PDC's social basis was rather cross-class [@rennwald_class_2014; @tawfik_les_2019]. Therefore, of the three right-wing government parties, only the Liberal-Radical party's social basis can truly be described as bourgeois. Regarding the support for the Socialist party, the latter was neither a bourgeois nor solely a popular party, but a hybrid-class party whose social basis was constituted by socio-cultural professionals and production workers [@rennwald_social_2020].

The coalition between the Radical-Democrats, Christian-Democrats and Agrarians against the left bloc took deeply roots into the Swiss socio-political system: it persisted during the post-war period and still shapes Swiss politics to some extent nowadays [@mazzoleni_voisinages_2013; @meuwly_les_2010; @meuwly_liberte_2008]. A striking feature of the Swiss DSB is its persistence and resilience since 1848: even though this bloc underwent important transformations, especially after the introduction of the proportional system in 1918, it was always mediated by right-wing parties, first the Radicals and then an extended coalition with the Catholic-Conservative and the Agrarians. Despite its incorporation in the parliament and government, the left bloc never truly became dominant in Switzerland.

To sum up, Switzerland's traditional social blocs crystallized into an opposition between two blocs, which is summarized in @tbl-socialblocs1. On the one hand, the dominant social bloc, which can be called liberal-conservative, gathered the large employers and owners of the big industrialized and internationalized firms, medium and small business owners and farmers, and skilled workers from the private sector. The main economic policy demands of the large industrialized firms were to maintain low social protection, low taxation, no barriers to external trade and to promote free trade at the international level. At the political level, this socio-political group was represented by the *Vorort* (now *EconomieSuisse*), a business interest associations, and the Radical-Democratic party. Alongside this group were the medium and small business owners producing essentially for the national market, and skilled workers (managers) from the private sector. The policy demands from the medium and small firms differed substantially from those of the large companies in terms of the degree of protection from international competition and public subsidies. Unlike the big corporations, small firms felt more threatened by external competition, and, since they produced mainly for the internal and rather small Swiss market, they tended to ask for protection and public subsidies.

+---------------------------+------------------------------------------------------------------+-----------------------------------------------+
|                           | Socio-political groups                                           | Policy demands                                |
+===========================+==================================================================+===============================================+
| Liberal-conservative bloc | -   Large industrial and internationalized firms                 | -   Low employment protection                 |
|                           |                                                                  |                                               |
|                           |                                                                  | -   Limited redistribution                    |
|                           |                                                                  |                                               |
|                           |                                                                  | -   Low tax level                             |
|                           |                                                                  |                                               |
|                           |                                                                  | -   Free-trade                                |
+---------------------------+------------------------------------------------------------------+-----------------------------------------------+
|                           | -   Medium and small firms, farmers                              | -   Low tax level                             |
|                           |                                                                  |                                               |
|                           |                                                                  | -   Public transfers (subsidies)              |
|                           |                                                                  |                                               |
|                           |                                                                  | -   Protection from international competition |
+---------------------------+------------------------------------------------------------------+-----------------------------------------------+
|                           | -   Skilled workers from the private sector                      | -   Competitiveness                           |
|                           |                                                                  |                                               |
|                           |                                                                  | -   Lower taxes                               |
+---------------------------+------------------------------------------------------------------+-----------------------------------------------+
| Left bloc                 | -   Socio-cultural professionals, workers from the public sector | -   State intervention                        |
|                           |                                                                  |                                               |
|                           |                                                                  | -   Increase in real wages                    |
|                           |                                                                  |                                               |
|                           |                                                                  | -   Expansionary macroeconomic policy         |
+---------------------------+------------------------------------------------------------------+-----------------------------------------------+
|                           | -   Production workers from the private sector                   | -   Increase in real wage                     |
|                           |                                                                  |                                               |
|                           |                                                                  | -   Social and employment protection          |
|                           |                                                                  |                                               |
|                           |                                                                  | -   Protection from external competition      |
+---------------------------+------------------------------------------------------------------+-----------------------------------------------+

: Social blocs in Switzerland during the 70s and 80s. Classification based on Rennwald (2014), Tawfiq (2019). {#tbl-socialblocs1}

On the other hand, the main social bloc excluded from the DSB was the left bloc, constituted by socio-cultural professional, workers from the public sector, and production workers from the private sector. These groups were mediated by the Swiss Socialist party, which was incorporated into the government during the Post-war period, but never really managed to dominate the parliament and government.

## Swiss social blocs and socio-economic factors: 1980-2016

How have the social bases of the Swiss left, center-right and radical-right evolved since the 70s? How are the support for the main Swiss social blocs influenced by socio-economic characteristics? The objective of this section is to explore the composition of the main Swiss social blocs in terms of socio-economic characteristics and assess their evolution since the post-war period. The previous chapter asked whether the periods of political instability measured through confidence in the Swiss government in the 1990s and early 2005 can be interpreted as a political crisis. By exploring the support for the main Swiss parties since 1981, this question can be partially answered, as well as the main research question of the present paper, which asks whether a dominant neoliberal Swiss social bloc is possible.

Moreover, a recent trend in political economy developed a tremendous analysis of the long run transformations of political cleavages, social blocs, and voting outcomes in almost all democratic countries around the world. The popularity of inequality studies, after the contributions of renown economists such as Atkinson [-@atkinson_inequality_2015], Milanovic [-@milanovic_global_2016] and of course Piketty [@piketty_capital_2014], economists recently became interested in the problematic of social conflict and political cleavages. One possible explanation of this sudden interest is the inequality paradox: increasing inequality and neoliberal reforms did not pave the way for growing support for redistribution, taxation of top income and wealth owners, or for left parties. The same paradox could be applied to a "climate change paradox", that is, the fact that the growing emergency or environmental issues were not followed by increasing support for green parties. For instance, the Swiss green parties were the biggest losers in the 2023 federal elections.

Those paradox thus gave way to a recent literature linking political cleavages and voting outcomes as function of a set of socio-economic factors, mainly income and education level. Trying to explore this inequality paradox, @piketty_brahmin_2018 underlines the fact that, rather than bringing back class-based divide back to its post-war salience, the global rise in inequality since the 1970s was followed by the emergence of a new cleavage and a multi-elite party system. During the post-war period, the electoral support for the left parties in The US, UK and France was mainly constituted by popular classes with low income and education levels whereas right-wing parties were supported by high income and education voters. The popular element of the support for the left blocs then transformed in the 1970-80s: left parties became the bloc of the highly educated, thus becoming the *brahmin left*, whereas the right kept the support of the wealthy (*merchant right*) [@piketty_brahmin_2018; @piketty_capital_2019]. A further project, which gave birth to the World Political Cleavage and Inequality Database (WPID), showed that these transformations affected almost all Western democracies [@gethin_political_2021].

Underlying the rise of a new cleavage in the 70s and 80s, Piketty and the WPID tackle a topic which is well-known in political science: the emergence of a cleavage centered around cultural values, that Piketty calls the "nativists-internationalists" divide, which cuts across the traditional class conflict and is the result of the increase in average education level and of globalization. This "cultural cleavage" has been given as many labels as it has analysts among political scientists: "materialist-postmaterialist" divide for Inglehart [-@inglehart_silent_1971; -@inglehart_value_1987; -@inglehart_culture_1990]; the "libertarian-authoritarian" cleavage [@kitschelt_transformation_1994; @flanagan_new_2003]; winners vs losers of globalization [@kriesi_globalization_2006; @kriesi_west_2008; @teney_winners_2014]; the "transnational cleavage" [@hooghe_cleavage_2018] or the "universalism--particularism" cleavage [@bornschier_cleavage_2010; @zollinger_cleavage_2022].

The combination of this new cultural cleavage with the traditional class-based conflict draws a bi-dimensional political space in which four social blocs are possible: the brahmin left, also called the internationalist-egalitarian bloc which is on the left both culturally and economically; an internationalist-inegalitarian bloc (merchant right) on the left culturally but economically on the right; a nativist-inegalitarian bloc on the right culturally and economically; and a nativist-egalitarian bloc which is culturally conservative but economically on the left [@piketty_brahmin_2018; @piketty_capital_2019].

Testing the relevance of the brahmin left vs merchant right divide, @amable_brahmin_2022 offer different perspectives over the possible coalitions resulting from this bi-dimensional political divide. On the one hand, preference for redistribution decreases with both education and income levels. On the other hand, support for globalization issues (such as immigration) is not only positively influenced by education level, but also positively associated with income. @amable_brahmin_2022 suggest the possibility of a rapprochement between the brahmin left and the merchant right over policy preferences, giving birth to a bourgeois bloc uniting the wealthiest and most educated groups from the left and the right. @amable_brahmin_2022-1 show that the brahmin left vs merchant right divide has not taken place in every country and that the classical popular left versus bourgeois right opposition is still relevant for many countries.

In this context of multi-elite party system and multidimensional political divides, Switzerland is an interesting case study. A widespread idea claims that the class-based cleavage always played a relatively minor role in Switzerland. In Switzerland, class conflict was always cut across a wide variety of other divides: religious, linguistic or regional. Moreover, due to strong standard-of-living increase throughout the 20th century, Switzerland is also often considered as a forerunner in the emergence of this "cultural" cleavage, and in the precociousness and strength of the educational shift.

@durrer_de_la_sota_party_2021 offer a detailed analysis of the links between socio-economic factors and the vote for the main Swiss social blocs. During the post-war period until the 1990s, the left bloc (mainly the PS) was supported by low income and education voters while right-wing parties showed different patterns of class support, which call into question the extent to which this bourgeois bloc can truly be characterized as "bourgeois". Of the three parties constituting the bourgeois bloc, only one of them, the PRD-PLR party, had clearly a bourgeois composition since it is consistently supported by the most wealthy and educated groups since the post-war period. The PDC and the UDC cannot really be described as "bourgeois" since, according to @durrer_de_la_sota_party_2021, the support for the two parties was negatively associated with income and education. While this negative association tends to decline for the PDC, it became even more strong for the UDC since the 90s. After this period, the left bloc became "brahmin" in the sense that it gathered strong support among the most educated classes, while its popular support moved either into abstention or to the UDC.

This article by @durrer_de_la_sota_party_2021, which is part of the WPID project, shows that Piketty's brahmin left vs merchant right divide is partially valid in Switzerland: a Swiss brahmin left indeed exists through the Swiss Socialist Party. However this bloc do not face a merchant right, but rather a bourgeois bloc gathered by the Radical-Liberals and a popular bloc which support the UDC or do not support any party. Rather than being constituted by a brahmin left, a merchant right and a nativist bloc, Switzerland's tripolar political space is rather divided between a bourgeois bloc, a brahmin left and a popular bloc.

+-----------------------------+------------------------------------------------------+------------------------------------------------------+
| **Income/**                 | **High level of income**                             | **Low level of income**                              |
|                             |                                                      |                                                      |
| **Education**               |                                                      |                                                      |
+=============================+======================================================+======================================================+
| **High level of education** | **Bloc Bourgeois**                                   | **Brahmin**                                          |
|                             |                                                      |                                                      |
|                             | <br>                                                 | <br>                                                 |
|                             |                                                      |                                                      |
|                             | -   PLR-PRD<br>                                      | -   PS since the 90s<br>                             |
|                             |                                                      |                                                      |
|                             | -   Greens since 2011-2019                           | <!-- -->                                             |
|                             |                                                      |                                                      |
|                             | <br>                                                 | -   Greens until 2011-2019<br>                       |
|                             |                                                      |                                                      |
|                             | <br>Policy preferences:                              | <br>Policy preferences:<br>                          |
|                             |                                                      |                                                      |
|                             | <br>                                                 | -   For redistribution<br>                           |
|                             |                                                      |                                                      |
|                             | -   Against redistribution<br>                       | <!-- -->                                             |
|                             |                                                      |                                                      |
|                             | -   Against state intervention<br>                   | -   For state intervention<br>                       |
|                             |                                                      |                                                      |
|                             | -   For equal rights & chances CH-foreigners         | -   For equal rights & chances CH-foreigner          |
+-----------------------------+------------------------------------------------------+------------------------------------------------------+
| **Low level of education**  | **Merchant**                                         | **Popular groups**                                   |
|                             |                                                      |                                                      |
|                             | <br>                                                 | <br>                                                 |
|                             |                                                      |                                                      |
|                             | -   None                                             | -   UDC<br>                                          |
|                             |                                                      |                                                      |
|                             | <br>                                                 | -   Abstention<br>                                   |
|                             |                                                      |                                                      |
|                             | <br>Policy preferences:<br>                          | -   PS until the 90s\<br                             |
|                             |                                                      |                                                      |
|                             | -   Against redistribution<br>                       | -   Christian parties until the 90s<br>              |
|                             |                                                      |                                                      |
|                             | -   Against state intervention<br>                   | <br>Policy preferences:<br>                          |
|                             |                                                      |                                                      |
|                             | -   For more rights and better chances for the Swiss | -   For redistribution<br>                           |
|                             |                                                      |                                                      |
|                             |                                                      | -   For state intervention<br>                       |
|                             |                                                      |                                                      |
|                             |                                                      | -   For more rights and better chances for the Swiss |
+-----------------------------+------------------------------------------------------+------------------------------------------------------+

: The four possible social blocs based on a bi-dimensional political divide and income and education levels. Party classification is based on Durrer de la Sota, Gethin, and Martinez-Toledano (2021) results (see their appendix). Policy preferences are based on Piketty (2018, 2019) and Amable and Darcillon (2022b) {#tbl-socialblocs2}

The transformation of party systems into a mutli-elite and tripolar political space came with important change in the structural basis of party support. @oesch_class_2010 show that, in Switzerland, the cultural cleavage is rooted in the class structure, and caused a re-alignment between class and party support. Before this shift took place, the traditional left bloc politically represented by the Socialist Party was a hybrid-class coalition gathering socio-cultural professionals (teachers, academics, journalists...) and production workers, but also from technicians, clerks and service workers [@rennwald_social_2020].

This strong class heterogeneity of the Swiss left declined: an analysis of Swiss post-election surveys for 2003 and 2007 show that the Swiss left was deserted by production workers who directed their support for the Swiss People's party due to cultural issues [@oesch_class_2010]. Using European Social Survey data for 2011 and 2015, @rennwald_social_2020 show that socio-cultural professionals remain the only "party preserve" of the Swiss left.

@oesch_electoral_2018 show that the rise of the radical-right and of the cultural cleavage transformed the patterns of class voting. The working class, or at least some of its constituencies such as production workers, are still attracted to the left due to their economic interests, but also by the far-right because of cultural issues, leaving socio-cultural professionals as the sole "stronghold" of the left. Small artisans and business owners are also divided between their support to the center-right and the far-right: their economic preferences make them closer to the center-right while their cultural preferences are closer to the far-right.

The empirical agenda of the present study is to test the validity and relevance of the literature reviewed above. More specifically, how are socio-economic factor shaping the composition of each Swiss social blocs? The objective of the empirical analysis conducted here is to test the following hypotheses regarding party support (PS), political leaning (PL), and issue positions (IP), derived from the contributions of the WPID project [@piketty_brahmin_2018; @piketty_capital_2019; @durrer_de_la_sota_party_2021; @gethin_political_2021; @gethin_brahmin_2022]; @oesch_electoral_2018; and Amable and Darcillon [-@amable_brahmin_2022; -@amable_brahmin_2022-1]:

1.  *Brahmin left versus merchant right hypotheses*

    -   PS1. The support for the left should be negatively associated with education and income levels in the first time periods. The negative association with education should then decrease and become positive over time.
    -   PS2. The support for the right should be positively associated with education and income in the first time periods and the positive association with education should decrease over time.

2.  *Bourgeois bloc hypothesis*

    -   The support for redistribution and state intervention should decrease with education and income. Support for equal rights between Swiss citizen and foreigners should increase with education and income.

### Data and empirical strategy

```{r import voxit}
#| echo: false
#| message: false
#| warning: false

vox <- read_sav("data/voxit.sav")
```

```{r create income and educ variable}
#| warning: false
#| message: false
#| echo: false

# Two ways to create the income variable
vox <- vox %>% 
  mutate(income = if_else(revenu == 1 & is.na(nivmena), 4,
                          if_else(revenu == 2 & is.na(nivmena), 3,
                                  if_else(revenu %in% c(3,4) & is.na(nivmena), 2,
                                          if_else(revenu == 5 & is.na(nivmena), 1,
                                                  if_else(is.na(revenu) & annee <= 1993, nivmena, NA))))))

         
vox$income <- ifelse(vox$income == 4, -1,
                    ifelse(vox$income == 3, 0,
                        ifelse(vox$income == 2, 1,
                          ifelse(vox$income == 1, 2, NA)))) # centering around zero, this will make the intercept more interpretable 


# for education, rescale the values so that the lowest level is equal to zero
vox <- vox %>% 
  mutate(education = case_when(
    educ == 1 ~ 0,
    educ == 2 ~ 1,
    educ == 3 ~ 2,
    educ == 4 ~ 3,
    educ == 5 ~ 4,
    educ == 6 ~ 5,
    is.na(educ) ~ NA
  ),
  educationf = case_when(
    educ == 1 ~ "comp_school",
    educ == 2 ~ "voc_training",
    educ == 3 ~ "maturity",
    educ == 4 ~ "higher_voctraining",
    educ %in% c(5,6) ~ "University",
    is.na(educ) ~ NA
  ),
  incomef = case_when(
    income == -1 ~ "low",
    income == 0 ~ "middle_low",
    income == 1 ~ "middle_high",
    income == 2 ~ "high"
  ))

vox$educationf <- relevel(factor(vox$educationf), ref = "voc_training")
vox$incomef <- relevel(factor(vox$incomef), ref = "low")
```

```{r party dummies and other recoding}
#| warning: false
#| message: false
#| echo: false

vox$pss <- ifelse(vox$p02 == 12, 1,
                  ifelse(is.na(vox$p02), NA, 0))

vox$udc <- ifelse(vox$p02 == 13, 1,
                  ifelse(is.na(vox$p02), NA, 0))

vox$plr <- ifelse(vox$p02 == 20, 1, ### le parti radical democratique fusionne avec le parti liberal en 2009
                  ifelse(vox$p02 == 4, 1,
                  ifelse(is.na(vox$p02), NA, 0)))

vox$green <- ifelse(vox$p02 == 6, 1,
                  ifelse(is.na(vox$p02), NA, 0))

vox$pdc_pcs <- if_else(vox$p02 %in% c(0, 1, 2), 1,
                       if_else(is.na(vox$p02), NA, 0))

vox$no_party <- if_else(vox$p02 %in% c(33), 1,
                       if_else(is.na(vox$p02), 1, 0))


vox$gender <- factor(vox$sexe, levels = c(1, 2), labels = c("male", "female"))
vox$married <- ifelse(vox$etatciv == 2, 1,
                      ifelse(is.na(vox$etatciv), NA, 0))

vox$regionL <- as_character(vox$regiling)
vox$religion <- relevel(factor(as_character(vox$confess)), ref = "Catholique romain")

## recoding age in 4 categories

vox$age_cat <- cut(vox$age, c(13, 30, 45, 65, 99), right = FALSE, include.lowest = TRUE)

vox$annee_cat <- cut(vox$annee, seq(from = 1981, to = 2016, by = 5), right = TRUE, include.lowest = TRUE)
vox$annee_f <- factor(vox$annee)

vox$canton <- as_character(vox$canton)


vox$house_ownership <- if_else(vox$statloge == 1, 1, 0)

### classes sociales dummies selon le secteur du répondant (posprof) ou du chef du menage (s31) sauf pour public
vox <- 
  vox %>% 
  mutate(
    farmer = if_else(posprof == 1 | s31 == 4, 1, 0), 
    self_employed = if_else(posprof == 2 | s31 == 1, 1, 0),
    ouvrier = if_else(posprof == 7 | s31 == 5, 1, 0),
    public = if_else(posprof == 6, 1, 0),
    atheist = if_else(confess == 4, 1, 0),
    protestant = if_else(confess == 1, 1, 0),
    catholic = if_else(confess %in% c(2, 5), 1, 0),
    lr = as.ordered(p04),
    
    left = if_else(p04 %in% c(0,1,2,3,4), 1,
                   if_else(is.na(p04), NA, 0)),
    center = if_else(p04 %in% c(5), 1,
                   if_else(is.na(p04), NA, 0)),
    right = if_else(p04 %in% c(6,7,8,9,10), 1,
                   if_else(is.na(p04), NA, 0)),
    stateint = if_else(val11 %in% c(1,2,3), 1,
                       if_else(is.na(val11), NA, 0)),
    redistribution = if_else(val5 %in% c(5,6,7), 1,
                       if_else(is.na(val5), NA, 0)),
    equal_rights_chforeign = if_else(val9 %in% c(1,2,3), 1,
                       if_else(is.na(val9), NA, 0))
    
  )

vox1 <- 
  vox %>% 
  filter(annee <= 1992) %>% 
  mutate(
    nivmenaf = factor(as_character(nivmena)),
    annee_cat = case_when(
      annee %in% c(1981:1985) ~ "[1981,1985]",
      annee %in% c(1986:1989) ~ "[1986,1989]",
      annee %in% c(1990:1992) ~ "[1990,1992]"
    )
  )

vox2 <- 
  vox %>% 
  filter(annee >= 1992) %>% 
  mutate(
    revenuf = factor(as_character(revenu)),
    income = case_when(
      revenu == 1 ~ "less_3k",
      revenu == 2 ~ "[3k,5k]",
      revenu == 3 ~ "[5k,7k]",
      revenu == 4 ~ "[7k,9k]",
      revenu == 5 ~ "more_9k"
      
    )
  )


vox2$income <- relevel(factor(vox2$income), ref = "less_3k")

vox2$annee_cat <- cut(vox2$annee, seq(from = 1992, to = 2016, by = 4), right = FALSE, include.lowest = TRUE)

```

To test the hypotheses above, two types of data sources are especially relevant. The *Swiss Election Studies (SELECTS) cumulative dataset* is a harmonized dataset merging every Swiss post-election (federal assembly elections, which take place every four years) surveys from 1971 to 2019 [@tresch_swiss_2022]. Since SELECTS data are already drastically analyzed by the literature (for instance by @durrer_de_la_sota_party_2021), I contribute to the existing research by analyzing another data source, the *VoxIt* cumulative dataset which harmonize every post-vote surveys from 1981 to 2016 [@brunner_enquetes_2017].

In each poll, respondents were asked which party they supported and their self-placement on the left-right axis on a Likert 10 points scale. The main dependent variables are dummy variables for party identification and political leaning computed from these two variables. Three other variables regarding issue position on public policies such as redistribution, state intervention and equal rights between Swiss citizen and foreigners are also considered.

The main independent variables are education and income levels. The variable "educ" is a qualitative variables on the highest achieved educational formation with 6 modalities ranging from primary school to university. It is important here to stress some peculiarities of the Swiss educational system. Switzerland's education system is strongly based on vocational training. As a result, other education categories tend to be missing for a lot of year-periods of the dataset. To solve this problem, but also to simplify the data structure, years are aggregated into several time periods to avoid separation issues.

```{r educational level plot}
#| message: false
#| warning: false
#| echo: false
#| label: fig-educ
#| fig-cap: "Education Levels in Switzerland: 1981-2016"
vox %>% 
  group_by(annee_cat) %>% 
  count(as_character(educ)) %>% 
  mutate(prop = n/sum(n)) %>% 
  drop_na() %>% 
  ungroup() %>% 
  ggplot()+
  aes(x = reorder(`as_character(educ)`, -prop), y = prop, fill = `as_character(educ)`)+
  geom_col()+
  facet_wrap(~annee_cat, strip.position = "top")+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.title = element_blank())+
  scale_fill_brewer(labels = c("Vocational Training", "Compulsory School", "Higher Voc Training", "Higher specialized school", "High school", "University"), palette = "Dark2")+
  scale_y_continuous(labels = scales::percent)+
  labs(title = "",
       x = "", y = "", caption = "Source: VoxIt")

```

@fig-educ shows the proportions of each level of education and their evolution over time. In the 1980s, most of the Swiss individuals completed either vocational training or compulsory school. The share of compulsory schooling declined over time, reflecting the overall increase in average education, as seen also by the increase in the share of university graduates. The share of vocational training declined over time, but is still the most widespread education level in Switzerland. For the regression analysis, the categories "university" and "higher specialized school" are merged together since the absence of the latter in the first two time periods could cause estimation issues. This does not pose a major issue since higher specialized schooling is rather similar to university in Switzerland.

Regarding income levels, the dataset contains two variables. The variable "nivmena" represents the houshold's standard of living in four categories (high, middle-high, middle-low, low) available from 1981 to 1991 whereas "revenu" is an income bracket variable in five categories available from 1992 to 2016. To harmonize these two variables in order to construct one single income or standard of living variable, I recode the variable "revenu" in order to make it similar to "nivmena" by grouping the income brackets categories 3 (from 5'001 chf to 7'000 chf) and 4 (from 7'001 to 9'000) into the "middle-high" category. This harmonized variable is used only in regressions for party support because political leaning and issue position variables are not available until 1992 (thus for these variables the variable "revenu" is used as an independent variable for income).

```{r income level plot}
#| message: false
#| warning: false
#| echo: false
#| label: fig-inc
#| fig-cap: "Descriptive statistics: variable nivmena and revenu"


plotincome1 <- 
vox %>% 
  filter(annee_cat %in% c("[1981,1986]", "(1986,1991]")) %>% 
  group_by(annee_cat) %>% 
  count(incomef) %>% 
  drop_na() %>% 
  mutate(prop = n/sum(n)) %>% 
  ungroup() %>% 
  ggplot()+
  aes(x = incomef, y = prop, group = incomef)+
  geom_col()+
  facet_wrap(~annee_cat)+
  scale_y_continuous(labels = scales::percent)+
  scale_x_discrete(limits = c("low", "middle_low", "middle_high", "high"))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  labs(y = "", x = "", caption = "", title = "Nivmena")

plotincome2 <- 
vox2 %>% 
  group_by(annee_cat) %>% 
  count(income) %>% 
  drop_na() %>% 
  mutate(prop = n/sum(n)) %>% 
  ungroup() %>% 
  ggplot()+
  aes(x = income, y = prop)+
  geom_col()+
  facet_wrap(~annee_cat)+
  scale_y_continuous(labels = scales::percent)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  labs(y = "", x = "", caption = "source: VoxIt", title = "Revenu")

cowplot::plot_grid(plotincome1, plotincome2)

```

A series of binary logit regressions for party support, political leaning, opinion on redistribution, state intervention and equal rights between Swiss and foreigners are conducted. The results presented in this section come from the following model:

$$ Y_{it} = \alpha + \beta_{I}I_{it} + \beta_{E}E_{it} + \beta I_{it}E_{it} + \gamma X_{it} + \eta_{1} Year + \eta_{2}YearI_{it}+ \eta_{3}YearE_{it} + \epsilon_{it} $$ {#eq-1}

With $I_{it}$ and $E_{it}$ the income and education level of individual $i$ in period $t$ and $X$ a vector of socio-demographic characteristics control variables which were available in the dataset for all time periods. The controls are for age (grouped into 4 categories), gender, religion, linguistic region (french, german, italian and romanche), marital status and a dummy for house ownership. $Year$ are dummy variables for the time periods and $\epsilon_{it}$ is the error term. The models include interaction terms between education and income as well as between income, education and time period in order to assess the temporal evolution of the correlation between the dependent and main explanatory variables. Income and Education are treated as categorical variables, with vocational training and low income as reference levels. The results presented below are essentially average marginal contrasts which allow to compare the difference in predicted probabilities between groups of interests, in our case, between highest and lowest income groups and between university graduates and vocational training graduates.

```{r plot ame parties}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ameparties
#| fig-cap: "Party support in Switzerland: 1981-2016. Average difference in  predicted probabilities between the highest and lowest income groups and between university and vocational training graduates."
#| fig-height: 5
ame_IP <- readRDS("C:/Users/celal/OneDrive/Bureau/master-thesis/data/ame/ame_IP.rds")
ame_parties <- readRDS("C:/Users/celal/OneDrive/Bureau/master-thesis/data/ame/ame_parties.rds")
ame_polleaning <- readRDS("C:/Users/celal/OneDrive/Bureau/master-thesis/data/ame/ame_polleaning.rds")

ame_parties %>% 
  filter(contrast %in% c("mean(high) - mean(low)", "mean(University) - mean(voc_training)")) %>%
  filter(party %in% c("ps", "greens", "noparty", "plr", "pdc", "udc")) %>% 
  ggplot()+
  aes(x = annee_cat, y = estimate, color = contrast, shape = contrast, group = party)+
  geom_point(position = position_dodge(width = 0.3))+
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), position = position_dodge(width = 0.3))+
  labs(x = "", y = "", caption = "Source: VoxIt")+
  theme_bw()+
  scale_y_continuous(labels=scales::percent)+
  geom_hline(yintercept = 0)+
  facet_wrap(~party)+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 50, vjust = 1, hjust=1))

```

@fig-ameparties show average marginal contrasts for each party or no party support in each time periods. The results partially confirm those of @durrer_de_la_sota_party_2021. First, the Swiss Socialist party (PS) used to be supported by relatively low education and income groups in the 1980s and became indeed a brahmin left party over time: the difference between the highly educated and vocational training graduates became positive in the second half of the 1980s and continued to increase in almost all time periods while the negative difference between the highest and lowest income groups decreased, indicating that the party's social basis became more educated and wealthier over time. However, the results show that the educational shift took place already in the 80s: university graduates already supported more the PS compared to vocational training graduates. This is an important result because it indicates that the PS became the party of the most educated in the 80s, before the economic and political crisis of the 90s and, most importantly, before the vote on the adhesion to the European Economic Area (EEA) in 1992, events which are often associated with the final stage of the educational shift and the PS's loss of popular support.

Second, the Liberal-Radical party (Radical-Democratic and Swiss Liberal Party before the latter merged in 2008) (PLR) shows a clear bourgeois basis for almost all time periods. The PLR is thus indeed the party of the wealthiest and most educated groups and is the only party that shows this pattern. However, the PLR seems to gradually become a merchant right party in the last time periods. In @durrer_de_la_sota_party_2021, another party, the Greens, also became bourgeois between 2011-2019, with the difference between the top 10% income and bottom 90% income voting Green becoming positive during this period while the difference between the top 10% and bottom 90% educated was positive since the 80s. The results shown in @fig-ameparties rather indicate that the Greens was a brahmin left party from 1981 to 2016.

Third, the results show that the Swiss People's Part (UDC) was a merchant right party whose social basis became popular both in terms of education and income levels. Before the 90s the party gathered the less educated, but also the wealthiest income groups (hence "merchant" in Piketty's terms). Here the results contrasts with those of @durrer_de_la_sota_party_2021 which show that the UDC was already a popular party gathering the less educated and poorest income groups in the 70s. Finally, we can see that popular groups, both in terms of education and income levels, are likely to not support any party relatively to the wealthy and most educated.

```{r plot ame political leaning}
#| echo: false
#| message: false
#| warning: false
#| label: fig-amepolleaning
#| fig-cap: "Political leaning in Switzerland: 1981-2016. Average difference in predicted probabilities between the highest and lowest income groups (less than 3000CHF per month vs more than 9000CHF), the second highest ([7'000chf-9'000chf]) and lowest income groups, and between university and vocational training graduates."
#| fig-height: 4

ame_polleaning$leaning[ame_polleaning$leaning == "center"] <- "don't know"
ame_polleaning %>% 
  filter(contrast %in% c("mean(University) - mean(voc_training)", "mean([7k,9k]) - mean(less_3k)", "mean(more_9k) - mean(less_3k)")) %>%
  ggplot()+
  aes(x = annee_cat, y = estimate, color = contrast, shape = contrast, group = leaning)+
  geom_point(position = position_dodge(width = 0.8))+
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), position = position_dodge(width = 0.8))+
  labs(x = "", y = "", caption = "Source: VoxIt")+
  theme_bw()+
  scale_y_continuous(labels=scales::percent)+
  geom_hline(yintercept = 0)+
  facet_wrap(~leaning)+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 50, vjust = 1, hjust=1))

```

@fig-amepolleaning show average marginal contrasts for left leaning, right leaning and no leaning. Interestingly, we can see that right leaning used to be bourgeois in the 80s and 90s while left leaning used to be brahmin. Over time, left leaning became bourgeois and right leaning became merchant. In the 80s and 90s, left leaning was positively associated with education, but not with income level while right leaning was positively associated with both. The positive difference between university and vocational training graduates declined for right leaning whereas it stayed rather stable for left leaning, indicating that the most educated support the left for all time periods while their support for the right declined over time. Popular groups tend to not identify with either the left or the right.

```{r plot ame issue position}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ameIP
#| fig-cap: "Opinion on several social issues in Switzerland: 1992-2016. Average difference in predicted probabilities between the highest and lowest income groups (less than 3000CHF per month vs more than 9000CHF) and between university and vocational training graduates."
#| fig-height: 4
ame_IP %>% 
  filter(contrast %in% c("mean(University) - mean(voc_training)", "mean(more_9k) - mean(less_3k)")) %>%
  ggplot()+
  aes(x = annee_cat, y = estimate, color = contrast, shape = contrast, group = issue)+
  geom_point()+
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high))+
  labs(x = "", y = "", caption = "Source: VoxIt")+
  theme_bw()+
  scale_y_continuous(labels=scales::percent)+
  geom_hline(yintercept = 0)+
  facet_wrap(~issue)+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 50, vjust = 1, hjust=1))
```

Finally, @fig-ameIP show average marginal contrasts for issue positions regarding equal rights between foreigner and Swiss citizen, redistribution and state intervention. The support for equal rights is positively associated with both education and income levels. However, the positive difference between highest and lowest income groups tend to decrease since the 2000s, indicating a decline of the "progressive attitude" of the wealthiest income groups towards foreigners. The difference between income groups in the support for redistribution is negative for all time periods whereas the difference between education groups was declining from 1992 to 2008, but tend to increase since in 2008-2012, but is not significant in 2012-2016. Regarding support for state intervention, the difference between education groups and income groups are respectively positive and negative: university graduates tend to be relatively more in favor of state intervention and the highest income groups tend to be relatively against for all time periods, indicating a brahmin left vs merchant right divide over state intervention.

To sum up, the results overall confirm the brahmin left vs merchant right hypotheses, even more than the results of the WPID project. The support for the Swiss Socialist party evolved from a traditional popular support towards a brahmin left support in the 80s. The results confirm the bourgeois composition of the PLR, since the latter seems to be gradually transforming into a merchant right composition since the beginning of the 2010s. The greens belong to the brahmin left since the 80s and there is no clear trend that the party evolved into a bourgeois party. Regarding the UDC, the latter used to be a merchant right party in the 80s, and evolved into a populist or, in Piketty's terms, a nativist party in the 90s. Thus, unlike @durrer_de_la_sota_party_2021 and the classification presented in @tbl-socialblocs, the social basis of the PLR seems to be transforming into a merchant right, while the Greens have kept their brahmin elements.

However, if one looks at political leaning on a left-right axis instead of party support, self-identification to the right used to be rather bourgeois in the 90s, and then went merchant afterwards. Self-identification to the left used to be brahmin in the 90s and even became bourgeois in recent time periods, if one looks at the difference in predicted probabilities between the upper-middle income and lowest income groups.

Finally, regarding policy preferences, preferences for equal opportunities and rights between foreigners and Swiss citizen for the highest incomes and the most educated seemed to converge until the mid-2000s, and have been diverging ever since. The divide over state intervention shows a brahmin vs merchant opposition in all time periods, with the highest income group being in favor of free market (relatively to the lowest income group) and university graduates being relatively more in favor of state intervention (relatively to vocational training graduates). The divide over income redistribution is less clear-cut than the state vs market divide: the cleavage appears to be brahmin vs merchant for some time periods (mostly 1992-1996 and 2008-20012), solely negatively correlated with income (1996-2004) or show a bourgeois vs anti-bourgeois opposition (2004-2008), with both the most education and wealthy being relatively opposed to redistribution.

Overall, the regression analysis conducted in this section, using a different data source from the WPID project and from Oesch and Rennwald, offer a new way to describe the evolution of the Swiss social blocs, which is summarized in table 2. However, this analysis suffers from several limitations.

On the one hand, one possible methodological issue of *VoxIt* dataset is the clustering structure of the data. Swiss citizen are clustered into several geographical levels, the most important ones being the cantonal and communal (district) levels. Switzerland is a highly decentralized and federalist country: every day political life as well as economic and political institutions can be substantially different across cantons. Every canton has its own party system and some Swiss parties, such as the *Lega Ticinessi* or *Mouvement Citoyen Genevois* (*Geneva Citizen Movement*) exist only at the cantonal level. A possible methodological solution, which would take into account both the time periods and the cantons as levels in which data are clustered, was considered by the author: a series of multilevel logit regressions. However, the results from these multilevel regressions showed that the variations between canton were negligible and thus a more simple model, the one presented in this section, was chosen. Moreover, the results between the multilevel model and the model of this section give overall similar results.

On the other hand, the perspective adopted in this section is essentially a top-down approach, in which social blocs are derived from the relationships between party support, political leaning or policy preferences taken as dependent variables to be explained by a set of socio-economic characteristics. This is the method which is the most widespread in the literature, but it suffers from having an oversimplified view of social conflict, which is almost always reduced to two dimensions (the so-called cultural and economic divides). One important contribution of the neorealist approach is to underline the fact that social conflict is multidimensional [@amable_multidimensional_2023] and that a bottom-up approach can be more suited to analyze the formation of social blocs. The next section will therefore adopt this bottom-up approach which is specific to the neorealist approach by conducting two latent class analysis for two different periods: 1999 and 2019.

## The transformation of the Swiss social blocs during the crisis of the 90s: between recomposition and resilience

However, since the break-up of the post-war Keynesian "historical compromise" between the traditional Swiss left and right-wing parties during the so-called "decade of all dangers" of the 90s [@boschetti_conquete_2007], this so-called bourgeois bloc is undergoing important transformations, that some authors interpreted as various forms of latent political crises. The Swiss political scene is still marked by the rise of the Swiss People Party (UDC), one of the most successful far-right populist parties in Europe. The electoral success of the UDC is often associated with the destabilization of the traditional bourgeois bloc, which entered a phase of "cacophonic" crisis [@meuwly_liberte_2008] which still persists in the early 2020s.

```{r party share graph}
#| warning: false
#| message: false
#| echo: false
#| fig-cap: "Party Strength: 1918-2019"
#| label: fig-partystrength
party_share <- read_excel(path = "data/force_parti_ofs.xlsx",
                          sheet = 2)
party_share[party_share == "*"] <- NA
party_share <- data.frame(lapply(party_share, as.numeric))

party_share_long <- 
party_share  %>% 
  pivot_longer(cols = !`...1`,
               names_to = "party",
               values_to = "Party_strength") %>% 
  rename(year = "...1")

party_share_long$party[party_share_long$party == "PLR..PRD..3."] <- "PRD-PLR"

party_share_long %>% 
  filter(party %in% c("PRD-PLR", "PDC", "PS", "UDC", "PES")) %>% 
  ggplot()+
  aes(x = year, y = Party_strength, color = party, shape = party)+
  scale_x_continuous()+
  geom_line()+
  geom_point()+
  theme_bw()+
  scale_color_brewer(palette = "Set1")+
  labs(caption = "Source: OFS", x = "", y = "Party Strength in %")

```

In fact, all the Swiss traditional main parties, the Swiss Socialist Party (PS) The Radical-Liberals (PLR) and the PDC, are in decline since the post-war period (see @fig-partystrength). On the right, the crisis can be seen through the fall in vote shares (apart from the SVP); and the increase in party merges and re-configurations. In 2008, the Bourgeois Democratic Party (PBD) was created after a scission from the SVP. One year later, the Radicals, who created the Swiss modern state in 1848 and dominated Swiss politics until the first half of the 20th century, merged with the liberals to create the PLR. The PBD, whose creation disturbed the composition of a reputed unchangeable federal council, then merged with the PDC to create *Le Centre* (the Center) in 2021. The federal election of 2023 led to major historical change for the Swiss right. The PLR, very far from its former hegemonic domination of Swiss politics, is now in the fourth place in terms of low chamber parliamentary seats.


