# Social blocs, socio-economic factors and voting outcomes in Switzerland

## The Swiss bourgeois and left blocs: historical perspective

Since the break-up of the post-war Keynesian "historical compromise" between the Swiss Socialist Party and the traditional right parties during the so-called "decade of all dangers" of the 90s [@boschetti_conquete_2007], the neoliberal transformation of the Swiss Hybrid Model was also followed by important political change, that some authors interpreted as various forms of latent political crises. The Swiss political scene is still marked by the rise of the Swiss People Party (SVP), one of the most successful far-right populist parties in Europe. The electoral success of the SVP is often associated with the destabilization of the traditional "bourgeois bloc", which entered a phase of "cacophonic" crisis [@meuwly_liberte_2008] which still persists in the early 2020s. 

```{r set up}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(haven)
library(quarto)
library(sjlabelled)
library(readr)
library(readxl)
library(haven)
library(viridis)
library(stargazer)
library(gtsummary)
library(gt)
library(brms)
library(ggrepel)
library(cowplot)
library(ggpubr)
library(lme4)
library(prediction)
library(marginaleffects)
library(ggeffects)
library(ordinal)
library(broom)
library(knitr)
library(readr)
library(gtExtras)
library(broom.mixed)
library(performance)
library(GGally)
library(questionr)
library(car)
library(effects)

```

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-cap: "Party Strength: 1918-2019"
#| label: fig-partystrength
party_share <- read_excel(path = "C:/Users/celal/OneDrive/Bureau/master-thesis/data/force_parti_ofs.xlsx",
                          sheet = 2)
party_share[party_share == "*"] <- NA
party_share <- data.frame(lapply(party_share, as.numeric))

party_share_long <- 
party_share  %>% 
  pivot_longer(cols = !`...1`,
               names_to = "party",
               values_to = "Party_strength") %>% 
  rename(year = "...1")

party_share_long %>% 
  filter(party %in% c("PLR..PRD..3.", "PDC", "PS", "UDC", "PES")) %>% 
  ggplot()+
  aes(x = year, y = Party_strength, color = party, shape = party)+
  scale_x_continuous()+
  geom_line()+
  geom_point()+
  theme_bw()+
  scale_color_brewer(palette = "Set1")

```

In fact, all the Swiss traditional main parties, the Swiss Socialist Party (PS) The Radical-Liberals (PLR) and the PDC, are in decline since the post-war period (see @fig-partystrength). On the right, the crisis can be seen through the fall in vote shares (apart from the SVP); and the increase in party merges and re-configurations. In 2008, the Bourgeois Democratic Party (PBD) was created after a scission from the SVP. One year later, the Radicals, who created the Swiss modern state in 1848 and dominated Swiss politics until the first half of the 20th century, merged with the liberals to create the PLR. The PBD, whose creation disturbed the composition of a reputed unchangeable federal council, then merged with the PDC to create *Le Centre* (the Center) in 2021. The federal election of 2023 led to major historical change for the Swiss right. The PLR, very far from its former hegemonic domination of Swiss politics, is now in the fourth place in terms of low chamber parliamentary seats. The domination of this bourgeois bloc constituted by the PLR, PDC and UDC, is a major characteristic of Switzerland's socio-political history and it is hence necessary to analyze its origins and transformations since the 19th century.

The combination of a religious and urban-rural divide between Protestants modernizers and Catholic federalists, which was historically the most structuring conflict in Switzerland, led to the Sonderbund civil war (1847) and the foundation of modern Switzerland in 1848. The federal institutions elaborated during and in the decades following the first Swiss constitution of 1848 can be interpreted as a social compromise between the losers of the Sonderbund civil war, the Catholic-Conservative coalition opposed to the centralization and modernization of the Swiss Confederation, and the victorious Radicals who were on average Protestants from urban areas [@sciarini_politique_2023]. However, these two blocs were highly heterogeneous and subject to internal contradictions.

From the outset of their movement, the Radicals were gathered by an unitary illusion: the pretention to transcend cleavages running through Swiss society and unite all its elements thanks to a higher goal: the modernization and centralization of the Swiss state. The "Radical kaleidoscope" was a mass-movement constituted by the working class through the Grütli Union and by an elite gathering entrepreneurs, bankers, journalists, lawyers and teachers, who were nonetheless divided over the role of the Swiss state. The democrats were partisans of a more centralized and interventionist state while the liberal or "manchesterian" Radicals, were more federalist and in favor of laissez-faire [@meuwly_les_2010]. 

Regarding the Catholic-Conservative bloc, the latter was from the start a cross-class coalition, constituted by, of course, the Catholic elite, but also by peasants, farmers, artisans and the commercial middle class from rural areas. These groups were united by Catholicism, but also by anti-modernism and federalism. This conservative coalition was the first to make extensive use of the constitutional and optional referendum (respectively 1848 and 1874) to block the modernization promulgated by the Radicals [@altermatt_conservatism_1979].

As industrialization, proletarianization and unionization intensified in the late 19th, the hegemony of the Radical mosaic gradually shattered. Part of the democrats and of the working class joined or even founded socialist organizations, as seen the creation of the Swiss socialist party in 1888 by a former Radical, Albert Steck. The creation of the latter and of the Radical-democratic party (PRD) in 1894 marked the beginning of the Radical shift towards liberal stances, which became clear-cut after the general strike of 1912, after which "a long-denied class conflict became evident" [@meuwly_les_2010, p.49]. Conversely, industrialization and class conflict broadened the social basis of the Catholic-Conservatives bloc towards the working class. The global development of a Catholic social doctrine, beginning with the encyclical *Rerum Novarum* (1981), was integrated by the Swiss Catholic-Conservatives, who then tried to attract parts of the working class [@altermatt_conservatism_1979].

This polarization around class conflict between a left and right bloc became even more salient after the second general strike of 1918, which was the biggest strike and social protest in Swiss history. The latter was followed by the introduction of the proportional representation system: the Radicals lost their hegemony in the parliament and the Christian democrats and the socialists became important political actors in the National Council. The socialist party (PS) consolidated a left bloc while the Radical-Democrats (PRD), the Catholic-Conservatives (PDC) and the Agrarians formed a liberal-conservative coalition. This coalition between the PRD, PDC and Agrarians and its close links with business association such as the Vorort (1870) led many observers to name this coalition the "bourgeois bloc" [@mach_redefinition_2007].

Until the post-war period, the evolution of the Swiss social blocs can thus be resumed as follows. From the middle to the late 19th century, the Radicals formed a strongly heterogeneous but dominant social bloc. This dominant social bloc was composed by the working class (mainly through the Grütli Union), and an urban elite composed by the most educated (journalists, lawyers, teachers) and the wealthy industrial bourgeoisie (entrepreneurs, bankers...). This DBS was united by their willingness to modernize the Confederation against the Catholic-Conservatives. A second period, from the late 19th to the second world war, marked the recomposition of the DBS and its transformation into a bourgeois bloc. Part of the working class and of the democrats (educated elite) joined the emerging socialist bloc while the Catholic-Conservatives attracted part of the working class while being progressively incorporated into the DBS, along with the Agrarian party which represented parts of the peasant and farmer classes. The coalition between the Radical-Democrats, Catholic-Conservatives and Agrarians against the left bloc took deeply roots into the Swiss socio-political system: it persisted during the post-war period and still shapes Swiss politics to some extent nowadays [@mazzoleni_voisinages_2013; @meuwly_les_2010; @meuwly_liberte_2008].

## Swiss social blocs and socio-economic factors: 1980-2016

What is the social basis of the Swiss left, center-right and radical-right in the last decades? How are the support for the main Swiss social blocs influenced by socio-economic characteristics? The objective of this chapter is to explore the composition of the main Swiss social blocs in terms of socio-economic characteristics and assess their evolution in the last decades. The previous chapter asked whether the periods of political instability measured through confidence in the Swiss government in the 1990s and early 2005 can be interpreted as a political crisis. By exploring the support for the main Swiss parties since 1981, this question can be partially answered, as well as the main research question of the present paper.

Moreover, a recent trend in political economy developed a tremendous analysis of the long run transformations of political cleavages, social blocs, and voting outcomes in almost all democratic countries around the world. The popularity of inequality studies, after the contributions of renown economists such as Atkinson [-@atkinson_inequality_2015], Milanovic [-@milanovic_global_2016] and of course Piketty [@piketty_capital_2014], economists recently became interested in the problematic of social conflict and political cleavages. One possible explanation of this sudden interest is the inequality paradox: increasing inequality and neoliberal reforms did not pave the way for growing support for redistribution, taxation of top income and wealth owners, or for left parties. The same paradox could be applied to a "climate change paradox", that is, the fact that the growing emergency or environmental issues were not followed by increasing support for green parties. For instance, the Swiss green parties were the biggest losers in the 2023 federal elections.

Those paradox thus gave way to a fresh and recent literature linking political cleavages and voting outcomes as function of a set of socio-economic factors, mainly income and education level. Trying to explore this inequality paradox, @piketty_brahmin_2018 underlines the fact that, rather than bringing back class-based divide back to its post-war salience, the global rise in inequality since the 1970s was followed by the emergence of a new cleavage and a multi-elite party system. During the post-war period, the electoral support for the left parties in The US, UK and France was mainly constituted by popular classes with low income and education levels whereas right-wing parties were supported by high income and education voters. The popular element of the support for the left blocs then transformed in the 1970-80s: left parties became the bloc of the highly educated while the right kept the support of the wealthiest. Piketty called the former bloc the "brahmin left" and the latter the "merchant right" [@piketty_brahmin_2018; @piketty_capital_2019]. A further project, which gave birth to the World Political Cleavage and Inequality Database (WPID), showed that these transformations affected almost all Western democracies [@gethin_political_2021].

Underlying the rise of a new cleavage in the 70s and 80s, Piketty and the WPID tackle a topic which is well-known in political science: the emergence of a cleavage centered around cultural values, that Piketty calls the "nativists-internationalists" divide, which cuts across the traditional class conflict and is the result of the increase in average education level and of globalization. The combination of this new cultural cleavage with the traditional class-based conflict draws a bi-dimensional political space in which four social blocs are possible: an internationalist-egalitarian bloc which is on the left both culturally and economically; an internationalist-inegalitarian bloc on the left culturally but economically on the right; a nativist-inegalitarian bloc on the right culturally and economically; and a nativist-egalitarian bloc which is culturally conservative but economically on the left [@piketty_brahmin_2018; @piketty_capital_2019]. This "cultural cleavage" has been given as many labels as it has analysts among political scientists: "materialist-postmaterialist" divide for Inglehart [-@inglehart_silent_1971; -@inglehart_value_1987; -@inglehart_culture_1990]; the "libertarian-authoritarian" cleavage [@kitschelt_transformation_1994; @flanagan_new_2003];
winners vs losers of globalization [@kriesi_globalization_2006; @kriesi_west_2008; @teney_winners_2014]; the "transnational cleavage" [@hooghe_cleavage_2018] or the "universalism–particularism" cleavage [@bornschier_cleavage_2010; @zollinger_cleavage_2022]. 

Testing the relevance of the brahmin left vs merchant right divide, @amable_brahmin_2022 offer contrasting results. On the one hand, preference for redistribution decreases with both education and income levels. On the other hand, support for globalization (such as immigration) is not only structured by education level, but also positively associated with income. @amable_brahmin_2022 suggest the possibility of a rapprochement between the brahmin left and the merchant right, giving birth to a bourgeois bloc uniting the wealthiest and most educated groups from the left and the right. @amable_brahmin_2022-1 show that the brahmin left vs merchant right divide has not taken place in every country and that the classical popular left versus bourgeois right opposition is still relevant for many countries.

In this context of multi-elite party system and multidimensional political divides, Switzerland is an interesting case study. A widespread idea claims that the class-based cleavage always played relatively a minor role in Switzerland. On the one hand, class conflict was always cut across a wide variety of other divides: religious, linguistic or regional. As a result, Switzerland was a forerunner in the emergence of this "cultural" cleavage, but also in the precociousness and strength of the educational shift. @durrer_de_la_sota_party_2021 offer a detailed analysis of the links between socio-economic factors and the vote for the main Swiss social blocs. During the post-war period until the 1990s, the left bloc (mainly the PS) was supported by low income and education voters while right-wing parties showed different patterns of class support, which call into question the extent to which this bourgeois bloc can truly be characterized as "bourgeois". Of the three parties constituting the bourgeois bloc, only one of them, the PRD-PLR party, had clearly a bourgeois composition since it is consistently supported by the most wealthy and educated groups since the post-war period. The PDC and the UDC cannot really be described as "bourgeois" since, according to @durrer_de_la_sota_party_2021, the support for the two parties was negatively associated with income and education. While this negative association tends to decline for the PDC, it became even more strong for the UDC since the 90s. After this period, the left bloc became "brahmin" in the sense that it gathered strong support among the most educated classes, while its popular support moved either into abstention or to the UDC.

The transformation of party systems into a mutli-elite and tripolar political space came with important change in the structural basis of party support. @oesch_electoral_2018 show that the rise of the radical-right and of the cultural cleavage transformed the patterns of class voting. The working class, or at least some of its constituencies such as production workers, are still attracted to the left due to their economic interests, but also by the far-right because of cultural issues, leaving socio-cultural professionals as the sole "stronghold" of the left. Small artisans and business owners are also divided between their support to the center-right and the far-right: their economic preferences make them closer to the center-right while their cultural preferences are closer to the far-right. 

The empirical agenda of the present study is to test the validity and relevance of the literature reviewed above. More specifically, how are socio-economic factor shaping the composition of each Swiss social blocs? The objective of the empirical analysis conducted here is to test the following hypotheses derived from the contributions of the WPID project [@piketty_brahmin_2018; @piketty_capital_2019; @durrer_de_la_sota_party_2021; @gethin_political_2021; @gethin_brahmin_2022]; @oesch_electoral_2018; and Amable and Darcillon [-@amable_brahmin_2022; -@amable_brahmin_2022-1]:

1. *Brahmin left versus merchant right hypothesis*

    The support for the left increases with education but not with income level. The support for right parties increase with income level, but not with education. The support for the far-right decreases with both income and education levels.
    
2. *Bourgeois bloc hypothesis*

    The support for right parties increase with both income and education levels. Popular groups support either the nativist far-right bloc or abstention.
    

### Data, empirical metholodogy 

One of the contributions of the present research is to draw extensive use the post-ballots Swiss surveys. In Switzerland, two types of political surveys are conducted: post-electoral and post-referendum surveys. The Swiss Elections Studies (SELECTS) conducts national surveys after each election of the national council while the VOX studies (now VOTO) conducts collects data after each referendum. 



